---
title: "Using RcppEnsmallen"
author: James Joseph Balamuta and Dirk Eddelbuettel
abstract: |
  Contained within are examples related to using RcppEnsmallen in everyday 
  work alongside of creating an _R_ package with it. 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RcppEnsmallen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`RcppEnsmallen` package provides an embedded copy of the
[`ensmallen`](https://www.ensmallen.org/docs.html) 
C++ library of optimization functions. Optimizers contained within are state
of the art and possess a high level of code quality. Each optimizer must be 
accessed through _C++_ by implementing the appropriate objective functions and,
then, surfaced into _R_ through
[`RcppArmadillo`](https://cran.r-project.org/package=RcppArmadillo). 
Alternatively, work has been done by Dirk Schumacher in 
[`armacmp`](https://github.com/dirkschumacher/armacmp) to automatically create
the underlying _C++_ code from _R_.  

**Note:** Optimizers in `RcppEnsmallen` only work with [`armadillo`](http://arma.sourceforge.net/docs.html)
data structures. Thus, if using [`Eigen`](http://eigen.tuxfamily.org/index.php?title=Main_Page) through 
[`RcppEigen`](https://cran.r-project.org/package=RcppEigen), please consider the 
[`RcppNumerical`](https://cran.r-project.org/package=RcppNumerical)
package.

## Linear Regression 

Consider the **Residual Sum of Squares**, also known as **RSS**, defined as:

$$RSS\left( \beta \right) = \left( { \mathbf{y} - \mathbf{X} \beta } \right)^{T} \left( \mathbf{y} - \mathbf{X} \beta \right)$$

The objective function we wish to minimize would be defined as:

$$f(\beta) = (y - X\beta)^2$$

The gradient is defined as:

$$\frac{\partial RSS}{\partial \beta} = -2 \mathbf{X}^{T} \left(\mathbf{y} - \mathbf{X} \beta \right)$$

### Two-Step Implementation

When using `ensmallen` to solve this problem, we must create a _C++_ class that 
computes both the objective function value and its gradient value either
together or separately under member functions named as:

- `Evaluate()`: Value of the objective function under the parameters.
- `Gradient()`: Convergence to the correct value under the given parameters. 
- `EvaluateWithGradient()`: Perform both steps at the same time. (Optional)

In the Linear Regression scenario, we will define each step separately to
emphasize the calculation occurring. Generally, the one step `EvaluateWithGradient()` 
function will be faster than the two step variant. More details on design can be
found on [`ensmallen` documentation page for differentiable functions](https://www.ensmallen.org/docs.html#differentiable-functions).

Before writing the class, `RcppEnsmallen` requires accessing the
library in a standalone C++ file with the follow include and Rcpp Attribute
declarations:

```{Rcpp header, eval = FALSE}
#include <RcppEnsmallen.h>
// [[Rcpp::depends(RcppEnsmallen)]]
```

The overaching Linear regression class should be constructed as follows: 

```{Rcpp classdef, eval = FALSE}
#include <RcppEnsmallen.h>
// [[Rcpp::depends(RcppEnsmallen)]]

// Define a differentiable objective function by implementing both Evaluate()
// and Gradient() separately.
class LinearRegressionFunction
{
public:
  // Construct the object with the given the design 
  // matrix and responses.
  LinearRegressionFunction(const arma::mat& X,
                           const arma::vec& y) :
  X(X), y(y) { }

  // Return the objective function for model parameters beta.
  double Evaluate(const arma::mat& beta)
  {
      return std::pow(arma::norm(y - X * beta), 2.0);
  }

  // Compute the gradient for model parameters beta
  void Gradient(const arma::mat& beta, arma::mat& g)
  {
      g = -2 * X.t() * (y - X * beta);
  }

private:
  // The design matrix.
  const arma::mat& X;
  // The responses to each data point.
  const arma::vec& y;
};
```

From there:

1. Construct a _C++_ function that exports into _R_.
2. Within the function, determine an appropriate optimizer for the problem. 
3. Combine the optimizer with the linear regression class to compute 
   the solution to the problem.

**Note:** Make sure to have the definition of the Linear Regression class in 
the same _C++_ file as the exported _C++_ function into _R_. 

```{Rcpp linreg, eval = FALSE}
// [[Rcpp::export]]
arma::mat lin_reg_lbfgs(const arma::mat& X, const arma::vec& y) {

  // Construct the first objective function.
  LinearRegressionFunction lrf(X, y);

  // Create the L_BFGS optimizer with default parameters.
  // The ens::L_BFGS type can be replaced with any ensmallen optimizer that can
  // handle differentiable functions.
  ens::L_BFGS lbfgs;

  lbfgs.MaxIterations() = 10;

  // Create a starting point for our optimization randomly.
  // The model has p parameters, so the shape is p x 1.
  arma::mat beta(X.n_cols, 1, arma::fill::randn);

  // Run the optimization
  lbfgs.Optimize(lrf, beta);

  return beta;
}
```

### Verifying Results

Prior to using the new optimizer in mission critical work, compare the results
to methods already implemented in _R_. The best way to achieve this is to create
an oracle model by specifying the parameters known to generate data and, then,
try to recover them. Moreover, if a method is already implemented in _R_ feel
free to try to check the result equality within an appropriate tolerance threshold.  

Following with this methodology, data must be generated. 

```{r data-generation}
n <- 1e6
beta <- c(-2, 1.5, 3, 8.2, 6.6)
p <- length(beta)
X <- cbind(1, matrix(rnorm(n), ncol = p - 1))
y <- X %*% beta + rnorm(n / (p - 1))
```

Next, the optimization procedure is used to estimate the parameters of interest.
Under this example, the results of the estimation can be compared to `lm()`. 
That said, `lm()` may have more precise results when compared against the 
optimizer as it is implemented with a closed-form solution to linear regression
plus the computational is performed more rigorously.  

```{r data-estimation, eval = FALSE}
coefs_lbfgs <- lin_reg_lbfgs(X, y)
coefs_lm <- lm.fit(X, y)$coefficients
```

```{r echo = FALSE}
coefs_lbfgs <- RcppEnsmallen::lin_reg_lbfgs(X, y)
coefs_lm <- lm.fit(X, y)$coefficients
compare_coefs = cbind(coefs_lbfgs, coefs_lm)
colnames(compare_coefs) = c("LBFGS", "LM")
rownames(compare_coefs) = paste0("Beta", seq_len(nrow(compare_coefs)))
knitr::kable(compare_coefs, longtable = FALSE, caption = "Comparison of Estimated Coefficients")
```

# Package Usage

RcppEnsmallen is best used within an _R_ package. The setup for `RcppEnsmallen`'s
use mirrors that of other `Rcpp`-based projects. In particular, the
`DESCRIPTION` file requires the `LinkingTo` field to contain:

```bash
LinkingTo: Rcpp, RcppArmadillo (>= 0.8.400.0.0), RcppEnsmallen (>= 0.2.10.0.1)
```

Next, the `src/` directory must contain both a `Makevars` and `Makevars.win` 
file with:

```bash
CXX_STD = CXX11
PKG_CXXFLAGS = -I. $(SHLIB_OPENMP_CXXFLAGS)
PKG_LIBS = $(SHLIB_OPENMP_CXXFLAGS) $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```

The `Makevars.win` file provides the appropriate configuration for Windows
while `Makevars` acts on Unix-alike systems like macOS, Linux, and Solaris.

